---
title: "Simulation and Optimisation in Sampling"
author: "Dr. Gianluca Boo, WorldPop, University of Southampton"
date: "2025-09-18"
format: revealjs
---

## Design-based vs. Model-based inference

- **Design-based inference**  
  - Validity relies on the randomization distribution given by the sampling design.  
  - No assumptions about the population model.  
  - Unbiasedness ensured if sampling probabilities are known.  

- **Model-based inference**  
  - Assumes a **statistical model** for the population values (e.g., regression, spatial).  
  - Efficiency comes from correctly specified models.  
  - Risks if the model is misspecified.  

ℹ️ Modern surveys often combine **design-based foundations** with **model-based enhancements**.

---

## Simulation-based evaluation

- Monte Carlo simulation allows evaluation of estimators under repeated sampling.  
- Used to assess **bias, variance, MSE** of different designs.  
- Steps:
  1. Define a population.  
  2. Repeat sampling many times.  
  3. Compute estimates each time.  
  4. Summarize performance.  

```{r}
#| echo: true
#| eval: false

set.seed(123)
library(tidyverse)

# Define artificial population
N <- 10000
pop <- tibble(id = 1:N, y = rlnorm(N, meanlog = 2, sdlog = 0.5))
mu_true <- mean(pop$y)

# Monte Carlo simulation for sample mean
n <- 200
R <- 1000
estimates <- replicate(R, {
  samp <- sample(pop$y, n)
  mean(samp)
})

mean(estimates)      # Average estimate (bias check)
var(estimates)       # Variance of estimator
mean((estimates - mu_true)^2) # MSE
```

---

## Calibration and variance estimation

- **Calibration**: Adjusts weights so sample totals align with known population totals.  
- Improves representativeness when auxiliary information is available.  

- **Variance estimation techniques**:  
  - **Jackknife**: systematically leave out one (or groups) of units.  
  - **Bootstrap**: resample with replacement to approximate sampling distribution.  

```{r}
#| echo: true
#| eval: false

library(survey)

# Example with jackknife and bootstrap variance
data(api)
dclus1 <- svydesign(id = ~dnum, weights = ~pw, data = apiclus1, fpc = ~fpc)

# Mean with jackknife
svymean(~api00, dclus1, na.rm=TRUE, vartype="se")

# Variance with bootstrap
boot_design <- as.svrepdesign(dclus1, type="bootstrap", replicates=50)
svymean(~api00, boot_design, na.rm=TRUE, vartype="se")
```

---

## Cost-effectiveness in sampling

- **Trade-off**: Precision ↑ vs. Cost ↑.  
- Survey design should balance both.  

- Sample size formula under budget:  
  \[
  n = \frac{B}{c}
  \]
  where  
  \(B\) = total budget,  
  \(c\) = unit cost per sampled element.  

- Optimal allocation often considers:  
  - Stratum costs  
  - Variance reduction per unit  
  - Fieldwork constraints  

```{r}
#| echo: true
#| eval: false

# Example: precision vs. cost
budget <- 10000   # total budget
cost_per_unit <- 50
n <- budget / cost_per_unit   # max sample size

sigma2 <- 25       # assumed population variance
se <- sqrt(sigma2 / n)

list(n = n, se = se)
```

---

## Key takeaways

- **Design vs. model inference** → different philosophies, often complementary.  
- **Simulation (Monte Carlo)** is essential to assess estimator performance.  
- **Calibration + resampling methods** improve accuracy and variance estimation.  
- **Cost-effectiveness analysis** ensures survey precision is achievable under real budgets.  
